Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-05-08 manuscript.tex
updated for zenodo of code

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
PlasmaDispersionRelations.jl, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}.

@@

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
PlasmaDispersionRelations.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of: calculating the
dispersion relation of plasmas consisting of an arbitrary number of species; and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are faster. It has been shown that it is possible to
qualitatively recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics\cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise due to 2D
linear physics effects but does not exclude the possiblity that nonlinear 1D
physics\cite{Chapman2018} may also play a part.
There are several important features missing from these calculations. Among them
are the effects of the spatial inhomogeniety of the magnetic field and background
plasma, and the quantisation associatied with toroidal periodicity. These
effects may create an eigenmode structure, not accounted for herein,
associated with the kinetic MCI.
Cottrell et al. \cite{Cottrell1993} does discuss the implications of propagation
angles on ICE signals and indicate that lower harmonics must propagate
increasingly close to exactly perpendicular. These effects remain the subject of
further work to incorporate them in linear and nonlinear studies of ICE.

@@

\bibitem{Dendy1995}
Dendy R O {\textit et al} 1995 \textit{Nucl. Fusion} \textbf{35} 1733–1742
%Dendy R O, McClements K G, Lashmore-Davies C, Cottrell G A, Majeski R, and Cauffman S 1995 \textit{Nucl. Fusion} \textbf{35} 1733–1742
%  , “Ion cyclotron emission due to collective instability of fusion prod-ucts and beam ions in TFTR and JET,”

@@

\bibitem{Cottrell1993} 
Cottrell G A {\textit et al} 1993 \textit{Nucl. Fusion} \textbf{33}, 31365–1387
%Cottrell G A, Bhatnagar V, Costa O D, Dendy R O, Jacquinot J, McClements K G, McCune D, Nave M, Smeulders P, and Start D 1993 \textit{Nucl. Fusion} \textbf{33}, 31365–1387
% 
% "Ion cyclotron emission measurements during JET deuterium-tritium experiments,"

@@

\bibitem{Ichimura2008}
Ichimura M {\textit et al} 2008 \textit{Nucl. Fusion} \textbf{48}, 035012
%Ichimura M, Higaki H, Kakimoto S, Yamaguchi Y, Nemoto K, Katano M, Ishikawa M, Moriyama S, and Suzuki T 2008 \textit{Nucl. Fusion} \textbf{48}, 035012
% , "Observation of spontaneously excited waves in the ion cyclotron frequency range on JT-60u,"

@@

\bibitem{Chapman2017}
Chapman B {\textit et al} 2017 \textit{Nucl. Fusion} \textbf{57} 124004
%Chapman B, Dendy R O, McClements K, Chapman S C, Yun G, Thatipamula S, and Kim M 2017 \textit{Nucl. Fusion} \textbf{57} 124004
% , "Sub-microsecond temporal evolution of edge density during edge localized modes in KSTAR tokamak plasmas inferred from ion cyclotron emission,"

\bibitem{Chapman2018}
Chapman B {\textit et al} 2018 \textit{Nucl. Fusion} \textbf{58} 096027
%Chapman B, Dendy R O, Chapman S C, McClements K G, Yun G S and Thatipamula S G 2018 \textit{Nucl. Fusion} \textbf{58} 096027

@@

\bibitem{McClements1996}
McClements K G {\textit et al} 1996 \textit{Phys. Plasmas} \textbf{3} 543
%McClements K G, Dendy R O, Lashmore-Davies C N, Cottrell G A, Cauffman S and Majeski R (1996) \textit{Phys. Plasmas} \textbf{3} 543

@@

\bibitem{Reman2019}
Reman B {\textit et al} 2019 \textit{Nucl. Fusion} \textbf{59} 096013
%Reman B, Dendy R O, Akiyama T, Chapman S C, Cook J W S, Igami  H, Inagaki S, Saito K, and Yun G, 2019 \textit{Nucl. Fusion} \textbf{59} 096013
% “Interpreting observations of ion cyclotron emission from large helical device plasmas with beam-injected ion populations,”

@@

\bibitem{Sharapov2014}
Sharapov S E {\textit et al} 2014 \textit{Phys. Plasmas} \textbf{21} 082501
%Sharapov S E, Lilley M K, Akers R, Ayed N B, Cecconello M, Cook J W S, Cunningham G, and Verwichte E 2014 \textit{Phys. Plasmas} \textbf{21} 082501
% , "Bi-directional Alfvén cyclotron instabilities in the mega-amp spherical tokamak"

@@

\bibitem{Heidbrink2006}
Heidbrink W W {\textit et al} A 2006 \textit{Nucl. Fusion} \textbf{46} 324
%Heidbrink W W, Fredrickson E, Gorelenkov N N, Rhodes T, Zeeland M A 2006 \textit{Nucl. Fusion} \textbf{46} 324
% Observation of compressional Alfvén eigenmodes (CAE) in a conventional tokamak.

@@

\bibitem{Fredrickson2019}
Fredrickson E D {\textit et al} 2019 \textit{Physics of Plasmas} \textbf{26}, 032111
%Fredrickson E D, Gorelenkov  N N, Bell R E, Diallo A, LeBlanc B P, Podestà  M, and NSTX Team 2019 \textit{Physics of Plasmas} \textbf{26}, 032111
% "Emission in the ion cyclotron range of frequencies (ICE) on NSTX and NSTX-U",

>>>>

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}.

@@

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
LinearMaxwellVlasov.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of: calculating the
dispersion relation of plasmas consisting of an arbitrary number of species; and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are faster. It has been shown that it is possible to
qualitatively recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics\cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise due to 2D
linear physics effects but does not exclude the possiblity that nonlinear 1D
physics\cite{Chapman2018} may also play a part.
There are several important features missing from these calculations. Among them
are the effects of the spatial inhomogeniety of the magnetic field and background
plasma, and the quantisation associatied with toroidal periodicity. These
effects may create an eigenmode structure, not accounted for herein,
associated with the kinetic MCI.
Cottrell et al. \cite{Cottrell1993} does discuss the implications of propagation
angles on ICE signals and indicate that lower harmonics must propagate
increasingly close to exactly perpendicular. These effects remain the subject of
further work to incorporate them in linear and nonlinear studies of ICE.

@@

\bibitem{Dendy1995}
  Dendy R O \textit{et al} 1995 \textit{Nucl. Fusion} \textbf{35} 1733–1742
%Dendy R O, McClements K G, Lashmore-Davies C, Cottrell G A, Majeski R, and Cauffman S 1995 \textit{Nucl. Fusion} \textbf{35} 1733–1742
%  , “Ion cyclotron emission due to collective instability of fusion prod-ucts and beam ions in TFTR and JET,”

\bibitem{LinearMaxwellVlasov.jl}
  Cook J W S, \textit{LinearMaxwellVlasov.jl (0.1.0)}, Zenodo. https://doi.org/10.5281/zenodo.6529288

@@

\bibitem{Cottrell1993} 
Cottrell G A \textit{et al} 1993 \textit{Nucl. Fusion} \textbf{33}, 31365–1387
%Cottrell G A, Bhatnagar V, Costa O D, Dendy R O, Jacquinot J, McClements K G, McCune D, Nave M, Smeulders P, and Start D 1993 \textit{Nucl. Fusion} \textbf{33}, 31365–1387
% 
% "Ion cyclotron emission measurements during JET deuterium-tritium experiments,"

@@

\bibitem{Ichimura2008}
Ichimura M \textit{et al} 2008 \textit{Nucl. Fusion} \textbf{48}, 035012
%Ichimura M, Higaki H, Kakimoto S, Yamaguchi Y, Nemoto K, Katano M, Ishikawa M, Moriyama S, and Suzuki T 2008 \textit{Nucl. Fusion} \textbf{48}, 035012
% , "Observation of spontaneously excited waves in the ion cyclotron frequency range on JT-60u,"

@@

\bibitem{Chapman2017}
Chapman B \textit{et al} 2017 \textit{Nucl. Fusion} \textbf{57} 124004
%Chapman B, Dendy R O, McClements K, Chapman S C, Yun G, Thatipamula S, and Kim M 2017 \textit{Nucl. Fusion} \textbf{57} 124004
% , "Sub-microsecond temporal evolution of edge density during edge localized modes in KSTAR tokamak plasmas inferred from ion cyclotron emission,"

\bibitem{Chapman2018}
Chapman B \textit{et al} 2018 \textit{Nucl. Fusion} \textbf{58} 096027
%Chapman B, Dendy R O, Chapman S C, McClements K G, Yun G S and Thatipamula S G 2018 \textit{Nucl. Fusion} \textbf{58} 096027

@@

\bibitem{McClements1996}
McClements K G \textit{et al} 1996 \textit{Phys. Plasmas} \textbf{3} 543
%McClements K G, Dendy R O, Lashmore-Davies C N, Cottrell G A, Cauffman S and Majeski R (1996) \textit{Phys. Plasmas} \textbf{3} 543

@@

\bibitem{Reman2019}
Reman B \textit{et al} 2019 \textit{Nucl. Fusion} \textbf{59} 096013
%Reman B, Dendy R O, Akiyama T, Chapman S C, Cook J W S, Igami  H, Inagaki S, Saito K, and Yun G, 2019 \textit{Nucl. Fusion} \textbf{59} 096013
% “Interpreting observations of ion cyclotron emission from large helical device plasmas with beam-injected ion populations,”

@@

\bibitem{Sharapov2014}
Sharapov S E \textit{et al} 2014 \textit{Phys. Plasmas} \textbf{21} 082501
%Sharapov S E, Lilley M K, Akers R, Ayed N B, Cecconello M, Cook J W S, Cunningham G, and Verwichte E 2014 \textit{Phys. Plasmas} \textbf{21} 082501
% , "Bi-directional Alfvén cyclotron instabilities in the mega-amp spherical tokamak"

@@

\bibitem{Heidbrink2006}
Heidbrink W W \textit{et al} A 2006 \textit{Nucl. Fusion} \textbf{46} 324
%Heidbrink W W, Fredrickson E, Gorelenkov N N, Rhodes T, Zeeland M A 2006 \textit{Nucl. Fusion} \textbf{46} 324
% Observation of compressional Alfvén eigenmodes (CAE) in a conventional tokamak.

@@

\bibitem{Fredrickson2019}
Fredrickson E D \textit{et al} 2019 \textit{Physics of Plasmas} \textbf{26}, 032111
%Fredrickson E D, Gorelenkov  N N, Bell R E, Diallo A, LeBlanc B P, Podestà  M, and NSTX Team 2019 \textit{Physics of Plasmas} \textbf{26}, 032111
% "Emission in the ion cyclotron range of frequencies (ICE) on NSTX and NSTX-U",
